---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

This Report is the first Assignment of two in the Coursera.org class "Reproducible Research"


## Loading and preprocessing the data

First we read the date from a .csv file using the following commands:
```{r}
setwd("H:/My Documents/Coursera/05 Reproducible Research")
data <- read.csv("activity.csv")
```

We also need to the following packages to run the R code in this Report:
```{r, message=FALSE}
library(sqldf)
library(lattice)
```


## What is mean total number of steps taken per day?

Now we will calculate the average steps per day and plot a Histogram:
```{r, message=FALSE}
result1 <- sqldf("select date, sum(steps) TotalSteps 
                  from data  
                  where steps is not null 
                  group by date")

hist(result1$TotalSteps, main = "Histogram of number of steps, w/o NA's", 
     xlab = "Number of steps per day")
```

The average number of steps taken per day are `r format(mean(result1$TotalSteps))` steps and median number of steps taken per day are `r median(result1$TotalSteps)` steps.


## What is the average daily activity pattern?

Now we want to see if there is a difference between different time interval during the day.  We calculate the average steps taken per every 5 minute interval and plot it:
```{r}
result <- sqldf("select interval, avg(steps) AverageSteps 
                 from data  
                 where steps is not null 
                 group by interval")

plot(result$interval, result$AverageSteps, type="l", main = "Average number of steps", 
     xlab = "Time of day (minutes 0 to 2355)", ylab = "Average number of steps")

```

The maximum number of steps were taken at `r result$interval[result$AverageSteps == max(result$AverageSteps)]` or `r paste( substr(sprintf("%04d", result$interval[result$AverageSteps == max(result$AverageSteps)]), 1, 2), substr(sprintf("%04d", result$interval[result$AverageSteps == max(result$AverageSteps)]), 3, 4), sep = ":")`


## Imputing missing values

The data has `r sum(is.na(data$steps))` rows of missing steps (NA's) so now we will add the average number of steps for the same time period to all rows where the number of steps is missing:
```{r}
data_wo_na <- sqldf("select case when a.steps is null 
                                then avg.AvgSteps
                                else a.steps
                            end steps,
                            a.date,
                            a.interval
                     from data a
                     join (select interval, avg(steps) AvgSteps 
                           from data  
                           where steps is not null 
                           group by interval) avg on avg.interval = a.interval
                    ")
```

Now our new dataset has `r sum(is.na(data_wo_na$steps))` rows of NA's.  Now we use our new dataset to calculate the total number of step taken per day and plot a Histogram:
```{r}
result <- sqldf("select date, sum(steps) TotalSteps 
                 from data_wo_na  
                 group by date")

hist(result$TotalSteps, main = "Histogram of number of steps, w/ imputed missing values)", 
     xlab = "Number of steps per day")
```

Now with the new dataset the average number of steps taken per day are `r format(mean(result$TotalSteps))` steps which is `r abs(mean(result$TotalSteps) - mean(result1$TotalSteps))`
`r if(mean(result$TotalSteps) > mean(result1$TotalSteps)){
    "steps greater than"
} else if(mean(result$TotalSteps) < mean(result1$TotalSteps)){
    "steps less than"
} else {
    "equal to"
}`
the one without the NA's.  The median number of steps taken per day are now `r median(result$TotalSteps)` steps which is `r abs(median(result$TotalSteps) - median(result1$TotalSteps))`
`r if(median(result$TotalSteps) > median(result1$TotalSteps)){
    "steps greater than"
} else if(median(result$TotalSteps) < median(result1$TotalSteps)){
    "steps less than"
} else {
    "equal to"
}`
the one without the NA's.


## Are there differences in activity patterns between weekdays and weekends?

Now we create a new variable dependant on the date variable, dayofweek, which is either "weekend" or "weekday". Then we calculate the average steps taken per 5 minute interval for weekend and weekday.  Finally we plot the result in two separate graphs:
```{r}
result <- sqldf("select date, interval, steps, 
                        case when strftime('%w', date) in ('0', '6') 
                            then 'weekend'
                            else 'weekday'
                        end dayofweek
                 from data_wo_na")

result2 <- sqldf("select dayofweek, interval, avg(steps) AvgSteps
                  from result
                  group by dayofweek, interval")

dayfactor <-factor(result$dayofweek) 

xyplot(AvgSteps ~ interval | dayofweek, data = result2, type = 'l', 
       main = "Average number of steps by weekend / weekday",
       xlab = "Time of day (minutes 0 to 2355)", ylab = "Average number of steps",
       layout = c(1, 2))
```
